iSEE: Interactive SummarizedExperiment Explorer

### 2018-04-26 manuscript/main.tex
Prepared sections to flesh out, and minor edits here and there.

Please list all authors that played a significant role in the research involved in the article. Please provide full affiliation information (including full institutional address, ZIP code and e-mail address) for all authors, and identify who is/are the corresponding author(s).

% TODO: full address and emails too?

@@

Please list up to eight keywords to help readers interested in your article find it more easily.

% TODO: pick up nice keywords!

@@

\section*{Introduction}
Interactive data exploration is critical to the analysis and comprehension of the data generated by high-throughput biological assays such as transcriptomics.
Exploration drives the formation of novel data-driven hypotheses prior to a more rigorous statistical analysis, and enables diagnosis of potential problems such as batch effects and low-quality samples.
To this end, visualization of the data in an intuitive and interactive interface is crucial, allowing researchers to examine the data from different perspectives across samples (e.g., experimental replicates, patients, single cells) and features (e.g., genes, transcripts, proteins, genomic regions).

@@

\section*{Description of the iSEE interface}
Here, we present the iSEE software package for interactive data exploration.
iSEE is implemented in R using the shiny framework \citep{chang2017shiny} and exploits data architectures from the open-source Bioconductor project \citep{gentleman2004bioconductor}.
The iSEE interface is initialized with a single function call accepting a SummarizedExperiment object as input \citep{huber2015orchestrating}.
This object stores one or more matrices of experimental observations as ``assays'', where columns and rows represent samples and features, respectively.
Per-feature or per-sample variables are stored in the ``rowData'' and ``colData'', respectively; this includes experimental metadata as well as any analysis results.
All of these data aspects can be directly examined in the multi-panel layout of the iSEE interface (Fig.~\ref{fig:iSEE}A).
The flexibility of the SummarizedExperiment class is the driving factor behind its broad deployment throughout the Bioconductor ecosystem, with uses in RNA sequencing \citep{love2014moderated}, methylation \citep{aryee2014minfi} and Hi-C analysis pipelines \citep{lun2016infrastructure}, amongst others.
By accepting a SummarizedExperiment object, iSEE immediately offers interactive visualization for a variety of data modalities, complementing the state-of-the-art analysis methodologies already available in R/Bioconductor packages.
%In particular, we leverage off the state-of-the-art allowing us to avoid re-implementing them in iSEE (as is commonly done in other standalone visualization tools).

A key feature of iSEE is the ability to dynamically transmit information between panels (Fig.~\ref{fig:iSEE}B).
Users can define arbitrary links between ``transmitting'' and ``receiving'' panels, whereby selections in transmitting tables and plots highlight the corresponding data points in receiving panels.
This facilitates exploration of the relationships between different aspects of the data -- for example, users can easily determine co-expression patterns of genes in a particular region of a reduced dimensionality embedding, by transmitting information from the reduced dimension plot to one or more feature expression plots.
The linking paradigm extends to multiple plots whereby a plot can transmit to multiple receivers, or a receiving plot can itself transmit to another plot.
This allows users to mimic the arbitrarily complex gating strategies often found in analyses of flow cytometry data \citep{finak2014opencyto}, with the advantage of being extensible to any (meta)data present in a SummarizedExperiment object.
Information can similarly be transferred between tables and plots, e.g., selecting a row determines how points are coloured in receiving plots.

iSEE also offers a host of other customizable features to improve data exploration.
Users can supply their own colour mappings to iSEE, to control how colour scales are chosen for each metadata variable and assay type across all panels in the application.
Row statistics tables can be augmented with dynamic annotation based on the selected row, linking to online resources such as Ensembl or Entrez. % do we need to cite them? at least ENSEMBL has updated citable papers
For large datasets, points can be downsampled in a density-dependent manner to accelerate rendering of the plots, improving the responsiveness of the interface without compromising the fidelity of the visualization.
After analyses are completed, users can deploy iSEE on a server with a bespoke step-by-step ``tour'' of their dataset, guiding the audience through an examination of the salient features in the data.
iSEE also automatically memorizes the exact R code that was used to generate every plot and the existing links,
extending previous work by \cite{marini2016interrepro}.
This allows users to easily reproduce the results of any exploratory analysis, simply by copying the code reported by iSEE into their own analysis scripts.

\section*{Conclusion}
iSEE provides a general interactive interface for visual exploration of a diverse range of `-omics datasets.
We provide a number of demonstrations at https://github.com/LTLA/iSEE2018, using iSEE to explore the TCGA RNA-seq collection \citep{piccolo2015TCGA}, the 4,000 PBMC single-cell RNA sequencing dataset from 10X Genomics \citep{zheng2017massively} and a published mass cytometry dataset consisting of over 170,000 cells \citep{bodenmiller2012cytof}.

>>>>

% Please list all authors that played a significant role in the research involved in the article. Please provide full affiliation information (including full institutional address, ZIP code and e-mail address) for all authors, and identify who is/are the corresponding author(s).
% TODO: full address and emails too? if so, in which format do we provide them?

@@

% Please list up to eight keywords to help readers interested in your article find it more easily.

Single-cell, gene expression, transcriptomics, proteomics, visualization, interactive, shiny, Bioconductor

% TODO: refine keywords

@@

\section*{Introduction}
Interactive data exploration is critical to the analysis and comprehension of the data generated by high-throughput biological assays such as transcriptomics.
Exploration drives the formation of novel data-driven hypotheses prior to a more rigorous statistical analysis, and enables diagnosis of potential problems such as batch effects and low-quality samples.
To this end, visualization of the data in an intuitive and interactive interface is crucial, allowing researchers to examine the data from different perspectives across samples (e.g., experimental replicates, patients, single cells) and features (e.g., genes, transcripts, proteins, genomic regions, microarray probe sets).

@@

\section*{iSEE supports multiple genomics data types}
Here, we present the iSEE software package for interactive data exploration.
iSEE is implemented in R using the shiny framework \citep{chang2017shiny} and exploits data architectures from the open-source Bioconductor project \citep{gentleman2004bioconductor}.
The iSEE interface is initialized with a single function call accepting a SummarizedExperiment object as input \citep{huber2015orchestrating}.
The SummarizedExperiment class stores one or more matrices of experimental observations as ``assays'', where columns and rows represent samples and features, respectively.
For instance, these may represent gene expression matrices of raw counts along with different normalization strategies.
In addition, per-feature or per-sample variables are stored in the ``rowData'' and ``colData'', respectively; those may include experimental metadata as well as any analysis results.
All of these data aspects can be readily examined alongside one another in the multi-panel layout of the iSEE interface (Fig.~\ref{fig:iSEE}A).

Notably, the flexibility of the SummarizedExperiment class is the driving factor behind its broad deployment throughout the Bioconductor ecosystem, with uses in RNA sequencing \citep{love2014moderated}, methylation \citep{aryee2014minfi} and Hi-C analysis pipelines \citep{lun2016infrastructure}, amongst others.
As a consequence, the widespread adoption of the SummarizedExperiment class and its specialized extensions such as the SingleCellExperiment class enables iSEE to offer immediate interactive visualization for a variety of data modalities, complementing the state-of-the-art analysis methodologies already available in R/Bioconductor packages.
%In particular, we leverage off the state-of-the-art allowing us to avoid re-implementing them in iSEE (as is commonly done in other standalone visualization tools).
% KRA: any other child class like SingleCellExperiment worth mentioning? e.g. the BSseq class

\section*{Description of the iSEE interface}

In this section, give a brief description of each panel type.

% the previous section mainly focuses on the input object

\section*{Dynamic linking between panels}

A key feature of iSEE is the ability to dynamically transmit information between panels (Fig.~\ref{fig:iSEE}B).
Users can define arbitrary links between ``transmitting'' and ``receiving'' panels, whereby selections in transmitting tables and plots highlight the corresponding data points in receiving panels.
This feature facilitates exploration of the relationships between different aspects of the data -- for example, users can easily determine co-expression patterns of genes in a particular region of a reduced dimensionality embedding, by transmitting information from the reduced dimension plot to one or more feature expression plots.
The linking paradigm extends to multiple plots whereby a plot can transmit to multiple receivers, or a receiving plot can itself transmit its own selection to another plot.
This allows users to mimic the arbitrarily complex gating strategies often found in analyses of flow cytometry data \citep{finak2014opencyto}, with the advantage of being extensible to any (meta)data present in a SummarizedExperiment object.
Information can similarly be transferred between tables and plots, e.g., selecting a row determines how points are coloured in receiving plots.

\section*{Advanced features and customization}

% KRA: Also, we should probably highlight somewhere that additional inputs can customize the startup state of the application

\subsection*{ExperimentColorMap: custom color maps}

iSEE also offers a host of other customizable features to improve data exploration.
Users can supply their own colour mappings to iSEE, to control how colour scales are chosen for each metadata variable and assay type across all panels in the application.
% TODO: KRA develop here

\subsection*{Linked feature annotations}

Row statistics tables can be augmented with dynamic annotation based on the selected row, linking to online resources such as Ensembl or Entrez. % do we need to cite them? at least ENSEMBL has updated citable papers

\subsection*{Density-dependent downsampling for speed}

For large datasets, points can be downsampled in a density-dependent manner to accelerate rendering of the plots, improving the responsiveness of the interface without compromising the fidelity of the visualization.

\subsection*{Code tracking and reproducibility}

After analyses are completed, users can deploy iSEE on a server with a bespoke step-by-step ``tour'' of their dataset, guiding the audience through an examination of the salient features in the data.
iSEE also automatically memorizes the exact R code that was used to generate every plot and the existing links,
extending previous work by \cite{marini2016interrepro}.
This allows users to easily reproduce the results of any exploratory analysis, simply by copying the code reported by iSEE into their own analysis scripts.

\section*{Demonstration data sets and tours}

\subsection*{Single-cell plate-based RNA-seq}
% TODO: better title, or do we drop this allen section altogether?

We demonstrate the application of iSEE to examine a subset of 379 cells from of mouse visual cortex \citep{tasic2016allen}.

% TODO: embed the allen demo data set

\subsection*{Single-cell droplet-based RNA-seq}
% TODO: better title, or do we drop the allen section altogether (above)?

We also demonstrate the use of iSEE to explore a larger data set of 4,000 PBMC single-cell RNA sequencing dataset from 10X Genomics \citep{zheng2017massively}.

% embed the allen demo data set

\subsection*{Bulk RNA-seq}

We also demonstrate how iSEE seamlessly scales up with larger data sets as the TCGA RNA-seq collection of measurements for 23,368 genes across 7,706 tumor samples \citep{piccolo2015TCGA}.

% TODO: embed the TCGA demo data set

\subsection*{CyTOF}

Lastly, we also showcase the use of iSEE to vizualise and refine gating a gating analysis, followed by the investigation of a functional marker between two sample groups, using a published mass cytometry (CyTOF) data set \citep{bodenmiller2012cytof}.

% TODO: embed the TCGA demo data set

\section*{Conclusion}
iSEE provides a general interactive interface for visual exploration of a diverse range of `-omics datasets.
We provide a number of demonstrations at https://github.com/LTLA/iSEE2018, using iSEE to explore the TCGA RNA-seq collection \citep{piccolo2015TCGA}, the 4,000 PBMC single-cell RNA sequencing dataset from 10X Genomics \citep{zheng2017massively} and a published mass cytometry dataset consisting of over 170,000 cells \citep{bodenmiller2012cytof}.
% TODO: repurpose the conclusion if we showcase the examples in earlier sections
